Spatial variation of the native colon microbiota in healthy adults

### 2017-11-09 submission/manuscript.tex
now with figs 1-4 updated
almost done with last pat edits

The microbiome has been implicated in the development of colorectal
cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
of these diseases vary along the axis of the digestive tract. Further,
variation in the structure of the gut microbiota has been associated
with both diseases. Here we profiled the microbiota of the healthy
proximal and distal mucosa and lumen to better understand how bacterial
populations vary along the colon. We used a two-colonoscope approach to
sample proximal and distal mucosal and luminal contents from the colons
of 20 healthy subjects that had not undergone any bowel preparation
procedure. The biopsies and home-collected stool were subjected to 16S
rRNA gene sequencing and Random Forest classification models were built
using taxa abundance and location to identify microbiota specific to
each site. The right mucosa and lumen had the most similar community
structures of the five sites we considered from each subject. The distal
mucosa had higher relative abundance of Finegoldia, Murdochiella,
Peptoniphilus, Porphyromonas and Anaerococcus. The proximal mucosa had
more of the genera Enterobacteriaceae, Bacteroides and Pseudomonas. The
classification model performed well when classifying mucosal samples
into proximal or distal sides (AUC=0.850). Separating proximal and
distal luminal samples proved more challenging (AUC=0.580) and specific
microbiota that differentiated the two were hard to identify. By
sampling the unprepped colon, we identified distinct bacterial
populations native to the proximal and distal sides. Further
investigation of these bacteria may elucidate if and how these groups
contribute to different disease processes on their respective sides of
the colon.

@@

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)YC calculator metrics in mothur version 1.39.5 (15).
\(\theta\)YC distances were calculated to determine the dissimilarity
between two samples. Random Forest classification models were built
using the AUCRF R package using a leave-one-subject out approach (17).
For each model the data was split into a 19-subject training set and a
1-subject test set. The model was built and cross-validated using
AUCRFcv on the training set. The model was then tested on the left-out
patient. This process was repeated iteratively for all subjects and
results plotted as Reciever Operator Characteristic curves using the
pROC R package (18). Resultant models were used to identify the OTUs
that were most important for classifying each location. Species-level
information for sequences of interest was obtained by aligning the
sequences to the GenBank nucleotide databse using blastn. The species
name was only used if the identity score was \(\ge\) 99\% over the
full-length of the contig and matched a single reference.

@@

Differences in community membership at the phyla level were tested using
the analysis of molecular variance (AMOVA) metric in mothur. Differences
in \(\theta\)YC distances by location were tested using the Wilcoxon
rank-sum test adjusted for multiple comparisons using the
Benjamini-Hochberg procedure.

>>>>

The microbiome has been implicated in the development of colorectal
cancer (CRC) and inflammatory bowel diseases (IBD). The specific traits
of these diseases vary along the axis of the digestive tract. Further,
variation in the structure of the gut microbiota has been associated
with both diseases. Here we profiled the microbiota of the healthy
proximal and distal mucosa and lumen to better understand how bacterial
populations vary along the colon. We used a two-colonoscope approach to
sample proximal and distal mucosal and luminal contents from the colons
of 20 healthy subjects that had not undergone any bowel preparation
procedure. The biopsies and home-collected stool were subjected to 16S
rRNA gene sequencing and Random Forest classification models were built
using taxa abundance and location to identify microbiota specific to
each site. The right mucosa and lumen had the most similar community
structures of the five sites we considered from each subject. The distal
mucosa had higher relative abundance of \emph{Finegoldia, Murdochiella,
Peptoniphilus, Porphyromonas} and \emph{Anaerococcus}. The proximal
mucosa had more of the genera \emph{Enterobacteriaceae, Bacteroides} and
\emph{Pseudomonas}. The classification model performed well when
classifying mucosal samples into proximal or distal sides (AUC=0.850).
Separating proximal and distal luminal samples proved more challenging
(AUC=0.580) and specific microbiota that differentiated the two were
hard to identify. By sampling the unprepped colon, we identified
distinct bacterial populations native to the proximal and distal sides.
Further investigation of these bacteria may elucidate if and how these
groups contribute to different disease processes on their respective
sides of the colon.

@@

Diversity analysis was performed using the Simpson diversity calculator
and \(\theta\)\textsubscript{YC} calculator metrics in mothur version
1.39.5 (15). \(\theta\)\textsubscript{YC} distances were calculated to
determine the dissimilarity between two samples. Random Forest
classification models were built using the AUCRF R package using a
leave-one-subject out approach (17). For each model the data was split
into a 19-subject training set and a 1-subject test set. The model was
built and cross-validated using AUCRFcv on the training set. The model
was then tested on the left-out patient. This process was repeated
iteratively for all subjects and results plotted as Reciever Operator
Characteristic curves using the pROC R package (18). Resultant models
were used to identify the OTUs that were most important for classifying
each location. Species-level information for sequences of interest was
obtained by aligning the sequences to the GenBank nucleotide databse
using blastn. The species name was only used if the identity score was
\(\ge\) 99\% over the full-length of the contig and matched a single
reference.

@@

Differences in community membership at the phyla level were tested using
the analysis of molecular variance (AMOVA) metric in mothur. Differences
in \(\theta\)\textsubscript{YC} distances by location were tested using
the Wilcoxon rank-sum test adjusted for multiple comparisons using the
Benjamini-Hochberg procedure.
